Complementary roles of Theta and Alpha in Multi-Item Associations

### 2019-07-15 supplemental.tex
splitting into separate supplemental

%Dear all,
%We ar writing to invite you to submit an article to an upcoming special issue of the journal Hippocampus with the tentative title: Computational models of the hippocampus and related structures.  The theme includes computational and theoretical models as well as theory-driven data analysis of the hippocampus and related cortical and sub-cortical structures. 
%
%Articles for the special issue will be due June 15, 2019 with a planned publication date in early 2020.  All articles will be published on-line when accepted, but we will need to be firm about the deadline in order to avoid delays in publishing the full issue with volume and page numbers.
% 
%
%The contributions to the special issue can be in all forms of manuscripts accepted by Hippocampus (Commentaries, Reviews, Articles, Rapid Communications).
% 
%
%As is common with these types of special issues, we ask that any contributing author commit to reviewing two articles for the special issue.
% 
%
%Please let me know by Dec. 15 if you plan to contribute to this special issue.  Please also send me a tentative title and potential author list (which can change later if necessary).
%
%Best regards,
%Michael Hasselmo
%and Ehren Newman

@@

\begin{document}
\title{Complementary roles of Theta and Alpha in Multi-Item Associations}
\author{Nicholas Ketz\textsuperscript{1}\footnote{Nicholas Ketz current affiliation is: Information and Systems Sciences Lab, HRL Laboratories, Malibu, CA 90265}, Tim Curran\textsuperscript{1} and Randall C. O'Reilly\textsuperscript{1}}
\date{
	\textbf{1} Department of Psychology and Neuroscience, University of Colorado Boulder  
	 \\
	\today
}

\maketitle

\section*{Acknowledgements}\label{sec:acknowledge}

\section*{Abstract}\label{sec:abstract}
%The Abstract should be clearly written in 300 words or less and should succinctly state the objectives of the study, experimental design, major observations and conclusions, and their major significance. The abstract should be intelligible to neuroscientists in general and should thus be free of specialized jargon and abbreviations. References should generally not be cited in the abstract, but if they are, the complete citation should be given. Please provide main keywords.
Through the use of neural network models of the hippocampus and surrounding cortex this work proposes a framework for understanding how time frequency signatures measured at the scalp can be used to track declarative memory processes, and make quantitative predictions about how information in stored in long term memory is altered by these processes.  The fundamental hypothesis explored here is that neural oscillations in the theta (3-8 Hz), alpha (8-12 Hz) frequency bands can be tied to specific neural mechanisms supporting declaritive memory, and that these oscillatory signatures can be tracked in human scalp EEG recordings to predict behavioral changes in declarative memory processes. Specifically that oscillatory power in the theta band positively correlates with the how much information the hippocampus is reactivating for a given retrieval event, and that power in the alpha band positively correlates with how much information is being inhibited from being retrieved.  We first explore this idea through computational simulations of thalamo-cortical-hippocampal interactions which make a series of predictions tested in a human EEG experiment using the Dependent Events paradigm.  Results largely support the idea that thalamically mediated changes in alpha power during episodic memory encoding and retrieval are related to the flow of information between the neocortex and hippocampus, and that the bandwidth of this information flow tracks with theta power.  Further, this model suggest the encoding a retrieval processes are inter-dependent and that attentional modulation of what gets retrieved from long term memory subsequently impacts the underlying representation that information is encoded into.

\section*{Keywords}
%Please provide up to 5-7 keywords. Keywords should be taken from those recommended by the US National Library of Medicine's Medical Subject Headings (MeSH) browser list at www.nlm.nih.gov/mesh.
Episodic Memory, Brain Waves, Neural Networks, Thalamus, Hippocampus

\section{Introduction}\label{sec:intro}

It is generally accepted that long-term memory encoding and retrieval are not isolated functions of any one brain area \cite{Gabrieli98,PrinceDaselaarCabeza05}.  However, there are many challenges in moving beyond the traditional studies focusing on individual regions  (e.g., concentrating on the prominent role of the hippocampus), to obtain insight into the nature of network-level dynamics.  An increasing body of work support the idea that oscillatory dynamics in multiple brain regions are associated with different aspects of  memory encoding and retrieval \cite{NyhusCurran10,HanslmayrStaudiglFellner12,Klimesch12}, %update refs here!!!
These studies have generally found a consistent relationship between several frequency bands and successful encoding or retrieval of experiences; namely that oscillatory power within the alpha (8 to 12Hz) and beta (13 to 30Hz) frequency bands generally decrease with these memory processes, while theta (3 to 8Hz) and gamma (30 to 100+Hz) increase in the regions directly involved in the memory operations. Why these relationships exist, and what they tell us about the larger functional organization of memory systems in the brain are still open questions.

One popular model suggests that thalamic circuits play a functional role in selectively gating communication between cortical regions by synchronizing them in various frequency bands \cite{ShermanGuillery06,SaalmannKastner11,KetzJensenOReilly15}.  These gating mechanisms, implemented  by oscillatory synchronization, are central to system level brain function and ultimately affect cognitive behavior such as memory encoding and retrieval.  Further, the particular oscillatory dynamics engaged by this selection process may provide indicators of which system-scale neural circuits are coordinating together to facilitate that particular form of memory processing.

A particular thalmo-cortical circuit of interest for memory processes involves the hippocampus and its associated subcortical network, including the anterior thalamus, and medial septum which are essential in the maintenance of theta oscillations (4 to 8 Hz) \cite{Buzsaki02}.  Theta oscillations have been shown to be critically involved in successful encoding and retrieval of episodic like memories in both humans and animals \cite{NyhusCurran10,JonesWilson05,FuentemillaBarnesDuzelEtAl14}, and it is believed to be primarily driven by the pacemaker cells within the medial septum.  The anterior thalamus, however,  provides the connectivity for the hippocampus to potentially modulate communication with the medial and orbital PFC through theta synchronization \cite{VanderWerfJollesWitterEtAl03,AggletonDumontWarburton11}.  Thus, the theta circuit is important for the core functioning of the hippocampus during encoding and retrieval.

%There is MEG evidence that theta synchronization with the hippocampus can facilitate Recollective memory processes in humans \cite{FuentemillaBarnesDuzelEtAl14}. There is some indirect MEG and BOLD evidence that the thalamus plays a role in this synchronization process \cite{FuentemillaBarnesDuzelEtAl14,PergolaRanftMathiasEtAl13}, which is bolstered by the known connectivity between the hippocampus and PFC through the anterior thalamus from rodents \cite{AggletonBrown99,AggletonDumontWarburton11}.  The thalamic gating model connects these ideas suggesting Recollective memory works through the hippocampus via theta synchronization, however to date there is no strong evidence supporting this claim in humans.

A complimentary thalamic circuit involves the pulvinar providing synchronization within the alpha (8 to 12 Hz) band between parietal and visual cortices with the medial, orbital and lateral PFC \cite{BarbasHenionDermon91,KievitKuypers77,LopesdaSilva91}.  This circuit is important for bottom-up stimulus-driven encoding processes in memory \cite{JutrasFriesBuffalo13}, but it can also inhibit contradictory or interfering information during internally guided retrieval \cite{ParkLeeKangEtAl14}.  Thus, its overall contribution to memory can be mixed, and depends on the nature of the task and what kinds of memory processing is required.

%The pulvinar has a well established role in coordinating alpha synchrony in animal models \cite{SaalmannKastner11,SaalmannKastner09,LopesdaSilva91,LopesdaSilvaVosMooibroekEtAl80}, and there is recent evidence in human models showing alpha oscillations provide an active function of suppressing cortex that would otherwise interfere with the memory retrieval \cite{WaldhauserJohanssonHanslmayr12,HanslmayrStaudiglAslanEtAl10,KhaderRosler11}.  There is also human fMRI evidence showing PFC modulation of visual cortex through the pulvinar \cite{SotoHumphreysRotshtein07,RotshteinSotoGrecucciEtAl11}.

Indeed, a recent spiking model of these oscillations called the Sync/deSync Model highlights the role of a synchronized hippocampus that accompanies a desynchronized neocortex during encoding and retrieval.  This model in particular focuses on the role of decreased alpha power related to successful encoding of new information \cite{ParishHanslmayrBowman18, HanslmayrStaresinaBowman16}.  Similarly, recent empirical studies of human intracranial recordings find evidence to support the complimentary roles of hippocampal synchrony and neocortical desynchrony in successful encoding and retrieval \cite{GriffithsMichelmannRouxEtAl18, StaresinaMichelmannBonnefondEtAl16}.  In general these results find a decrease in neocortical alpha power and an increase in hippocampal gamma power that coincide with successful encoding and retrieval of declarative memories.

This accumulated evidence suggests a relationship between theta and alpha frequency bands and long-term memory systems as constructed by the Complimentary Learning Systems theory \cite{McClellandMcNaughtonOReilly95,OReillyBhattacharyyaHowardEtAl14}.  Namely, that theta power positively tracks successful engagement (encoding/retrieval) between hippocampus and neocortex, while alpha power negatively tracks with this process as active inhibition of cortex is released in response to the exchange of information between hippocampus and neocortex.  This relationship is further supported by the connectivity of the thalamus with cortex such that a specific model for how information is routed into and out of these systems can be tied to empirical measures of neural oscillations.  This work seeks to test this relationship through a series of computational simulations of hippocampal-thalamo-cortical interaction, and an accompanying EEG study in humans to compare against simulations results.

\subsection{Dependent Events Paradigm}
%
%Hippocampal dependent memory is generally thought to be associative in nature, where multiple cortical representations are bound together in a distinct hippocampal representation during encoding.  Later, an incomplete set of those cortical representations can be used as a cue to reactive the full set during retrieval \cite{McClellandMcNaughtonOReillyEtAl92}.  This multi-element type of memory retrieval is also referred to behaviorally as `Recollection' where multiple details of a given experience can be retrieved, and hippocampal activity increases with subjective confidence and the amount of information being retrieved \cite{RuggVilbergMattsonEtAl12}.  
%
%In contrast to Recollection, memories that exhibit `Familiarity' have no measurable details that can be recalled, however individual elements can be successfully recognized when presented with a mix of lures.  This Familiarity process is thought to rely more heavily on the medial temporal lobe (MTL) cortex surrounding the hippocampus, including entorhrinal, parahippocampal, and perirhinal cortices.  These cortical areas, among others, are proposed to be a separate memory system from the hippocampus, serving behavior when recall of all associated details is not necessary or available \cite{EichenbaumYonelinasRanganath07,AggletonBrown99}.  This theory of separate memory systems has been contentious for several decades with evidence both in support and in contrast to its core ideas  \cite{EichenbaumYonelinasRanganath07,WixtedSquire10}; namely that Recollection can occur in addition to Familiarity and that the hippocampus is not specific to Recollection but can be used in Familiarity based processing.  What is clear is that MTL cortex learns and maintains representations required for hippocampal retrieval, and those MTL regions along with the hippocampus can be used to re-instantiate cortical representations outside MLT during retrieval \cite{RuggVilberg13}.  
%

\begin{figure}
    \begin{center}
	\includegraphics[width=.9\textwidth]{figs/deev.pdf}
    \end{center}
      \caption[Dependent Events Paradigm]{
       Dependent events behavioral paradigm, taken from \textcite{HornerBisbyBushEtAl15} (a) Encoding phase. Participants see multiple paired associates and imagine each pair `interacting in a meaningful way as vividly as possible' for 6 s. Solid lines and dotted lines were not present, but highlight a closed-loop (solid lines) and open-loop (dotted lines) event. (b) Retrieval phase.  Participants were presented with a single cue and required to retrieve one of the other elements from the same event from among five lures within 6 s. This was followed by a 1-4 confidence rating within 6 s.  All 6 possible cue-target combinations are tested (c) The associative structure of closed-loop events, with example encoding order for the three pairwise associations (numbers 1-3). (d) The associative structure of open-loop events, with example encoding order for the three pairwise associations.
	}
     \label{fig:deev}
\end{figure}

To test this relationship we choose a recent hippocampal dependent multi-item association study to base our computational and empirical investigations on \cite{HornerBisbyBushEtAl15}.  In this study the authors used a previously developed paradigm (see \textcite{HornerBurgess13,HornerBurgess14}) in which they manipulated the binding of a set of overlapping paired associates.   Here two types of associative stimuli were studied, closed-loop and open-loop. In the `closed-loop' events participants studied a set of paired associates on separate trials, e.g. Person-Location, Location-Object, and Object-Person, where each category refers to the same stimulus for a given event.  This created a complete set of binding between all elements.   In contrast, the `open-loop' events incorporated a new element and left some relationships unstudied, e.g. Object-Location, Location-Person, Person-Animal.  Each of the associations was later tested in a cued recognition task, which showed greater inter-event Dependency for the closed-loop compared to the open-loop associates.  Dependency here is operationalized as the retrieval of \emph{all} associated pairs from a given event, or none; whereas no Dependency would imply retrieval of some of the associated pairs but not all.  The goal of this metric is to capture the strength of the binding between elements within a given event, where high Dependency indicates strong associations leading to retrieval of all items within a given event, and low Dependency indicates weak associations leading to some of the items being retrieved incorrectly. In contrast, a non-associative measure of accuracy can be made using an Independent model which calculates Dependency based on the assumption that each retrieval trial is statistically independent from the next, see Section \ref{sec:CalcDep} for more details.  %This provides a nice measure of both Recollection yielding high Dependency from strong associations between elements, but also uses a recognition paradigm where successful retrieval with low between element Dependency implies a similar construct to Familiarity, i.e. retrieval without detailed recall.

%theta increase a marker of hippocampal retrieval (Nyhus and Curran, 2010), and should correlate with amount of cortex reinstated
In this paradigm, closed-loop events show significantly greater Dependency as compared to the Independent model while the open-loop events showed no difference compared to its Independent model, implying that closed-loop events have a stronger between element binding as compared to open-loop events.  The closed-loop condition also showed more cortical reactivation of non-target elements within an event (i.e. studied elements not currently the target of the recognition trial), and that this reactivation correlated with hippocampal activity.  Results from fMRI studies such as \textcite{HornerBisbyBushEtAl15} provide spatially specific evidence for the hippocampus driving cortex to reinstate previous studied elements, but provide no evidence relating these processes to potentially related neural oscillations.  A computational model tied to the neural oscillations of this phenomena can provide an opportunity to bring clarity to this relationship by testing if patterns in theta power can be predicted by a computational model of the hippocampus.  Key aspects, however, relating to the focus of attention and alpha power are required to adequately model the paired associate and cued recalled components of the Dependent Events paradigm.

\subsection{Theta-Phase Model}

Building off our previous connectionist model of the hippocampus \cite{KetzMorkondaOReilly13}, referred to here as the Theta-Phase model, we seek to incorporate the ability to attend to particular dimensions of a given input pattern, while not attending to others.  Practically, this requires that each input be described on all possible stimulus dimensions, otherwise the hippocampal model will attempt to `recall' the missing information for those missing dimensions.  For example, one valid cortical representation experienced will contain information about size and shape but never in color, while another equally valid representation will contain information about color and shape, but never size.  The neural substrate supporting these representations needs to be able to represent all three dimension (and potentially many others) simultaneously and independently, but also be able to understand the inherent bias in which dimensions are likely to covary together.

%This problem begins to touch on the issue of intentional vs. incidental retrieval.  Intentional retrieval implies a goal directed, attentionally mediated re-activation of previously experienced representations, while incidental retrieval implies the reactivation of representations that are likely to be experienced together.  The Theta-Phase model is built exclusively on incidental retrieval mechanisms and currently has no way of reflecting the attentional, or goal directed, focus of the retrieval process.  Similarly, the lack of an intentional component to the retrieval process requires the Theta-Phase model to have valid, useful information in all potential stimulus dimensions, which is simply not possible in every scenario.  However, it is also crucial to maintain this incidental component to the model as it is fundamental to the Complimentary Learning Systems theory, and the main mechanism behind pattern completion \cite{McClellandMcNaughtonOReillyEtAl92,NormanOReilly03}.  What is required is an ability to balance between intentional and incidental retrieval drives in such a way that fits with known behavioral patterns.

A simple and ecologically valid way to reconcile this discrepancy is a mechanism that would highlight which of the stimulus dimensions are informative, or relevant given the current input.  This would be very similar to a selective attention mechanism.  Building from the evidence in support of an alpha mediated thalamic gating mechanism, we will explore modulation of cortical activity via thalamo-cortical interactions in the alpha band as a mechanism for the Theta-Phase model to gate specific stimulus dimensions for targeted encoding and retrieval.  The behavioral and physiological consequences of this mechanism will then be explored in a long-term memory paradigm that utilizes multi-element traces, and a bottom-up attentional signal to focus on specific stimulus dimensions.

\subsubsection{Attention and Alpha Oscillations}
\textcite{ShermanGuillery06} provide the most complete framework for how the flow of information can be guided from low level sensory regions into higher level cortical regions best suited for task specific processing.  The simple yet compelling idea is that cortical-cortical projections convey the contentful output from a given cortical region, while the thalamo-cortical projections contain the gating information.  Sensory information is allowed to cascade from one processing region to the next through cortical-cortical projections, provided that a selective thalamo-cortical gating signal alerts/allows the next cortical region to accept (or block) the incoming cortical-cortical signal.

The other critical aspect of this bottom-up attention mechanism is the connection to alpha oscillations.  Historically, alpha power has been shown to correlate with the onset of sleep, which led to the suggestion that high alpha power indicates an idle or perhaps inhibited cortical region.  The mechanistic connection between alpha power and thalamic gating is capture in the firing properties of the thalamic relay cells (TRC) which show two distinct modes of firing.  The first shows oscillatory firing patterns acting like a pacemaker system in the alpha frequency range (approximately 10 Hz or 100ms periodicity), and is generally referred to as bursting.  This bursting mode is believed to be engage when a given cortical region is in an idle state, i.e. when no active processing is occurring in the thalamus or the corresponding cortex linked to that thalamic relay region.  The other regime, referred to as tonic firing, shows no specific oscillatory firing pattern, and instead shows increased spontaneous firing in both the thalamus and corresponding cortical regions \cite{LerescheLightowlerSolteszEtAl91,ShermanGuillery06,LopesdaSilva91}.  These dynamics have been studied extensively as a naturally occurring phenomena of thalamic interactions with cortex, however simulation studies have also shown the potential for these mechanisms to be used in a goal directed way \cite{HindriksvanPutten13,VijayanKopell12,LeeWhittingtonKopell13}. The implications of these results are that cortical regions being driving by TRCs in bust mode will show more alpha power, and will process the incoming cortical-cortical information less veridically, while cortical regions being driven by tonic TRC firing will show a relative decrease in alpha power and process cortical-cortical information more completely.  This is the foundation for the connection between alpha power and thalamic gating, such that increased alpha power is indicative of decreased cortical processing.

The detailed biology of this framework is well captured by \textcite{ShermanGuillery06}, and relies on two critical components not yet implemented in the Theta-Phase model: 1. The thalamic relay cells (TRC) that pass the gating signal onto the next cortical region, and 2. The cortical neurons in layers 5 and 6 that send this gating signal to the next TRC, and trigger the opening of the gate for processing in associated cortex for that TRC.  The following sections provides a high-level description of how these components are implemented, while supplementary info section \ref{appDeepLeabra} illustrates the detailed implementation into the LEABRA framework.

\begin{figure}
	\includegraphics[width=\textwidth]{figs/depNet.pdf}
        \caption[Thalamic Gating as Implemented in Theta-Phase Model] {  \emph{(A)} Schematic of Theta-Phase model with thalamic gating.  Posterior cortex connects with both layers 2/3 of Entorhinal cortex (EC) via cortical-cortical projects, but also to thalamic relay cells (TRC) via layer 5 bursting cells.  These TRC units maintain their topography into EC layer 4 where they provide a gating signal for EC to accept or suppress the corresponding cortical-cortical projection. Topographic connectivity shown in dashed arrows, and distributed connectivity shown in solid.  \emph{(B)}  Neural network showing TRC implementation as single units for a given cortical area's input and corresponding unit group in EC.  This topography is kept consistent between TRC and EC unit groups.  \emph{(C)} Conceptualization of EC unit groups to paired associate elements and thalamic gating signal manifested in EC with 2 bottom-up cortical inputs missing (i.e. object and animal) in unit groups with a low (red) gating signal, and 6 inputs present (i.e. location, person, and 4 contextual elements) shown with a high (yellow) gating signal.  The cortical input topography is maintained between EC and TRC where TRC activation is used to calculate the normalized gating signal shown in the corresponding EC unit groups.  Note the gating signal is not representation specific, but rather generalized for a given unit group. 
        }
     \label{fig:depNet}
\end{figure}

\section{Computational Methods}\label{sec:compmethods}

In this work we start with the Theta-Phase model of the hippocampus and entorhinal cortex, as described in \textcite{KetzMorkondaOReilly13} and implemented in the Emergent framework \cite{AisaMingusOReilly08}.  Briefly, this model is built upon a series of structural and functional hypotheses based on anatomical and physiological data, which have been captured in the complementary learning systems (CLS) model of the hippocampus \cite{McClellandMcNaughtonOReilly95, NormanOReilly03}. The Entorhinal Cortex (EC) in the model is assumed to be the cortical gateway to the hippocampus.  This gateway feeds through the tri-synaptic pathway to the Dentate Gyrus (DG), CA3, and then to CA1.  Similarly, there is a parallel connection through the mono-synaptic pathway from the EC to the CA1 (and back).

The tri-synaptic pathway connections are broadly diffuse, and support the conjunctive binding of various distributed pieces of information into an overall episodic memory representation in the CA3. The mono-synaptic pathway conversely is topologically organized, not diffuse.  We capture this connectivity by organizing the simulated neurons in EC and CA1 into mutually interconnected {\em slots}, presumably encoding different separable elements/stimulus dimensions that loosely correspond to independent cortical regions converging on the EC \cite{Witter10}.  This slot architecture enables the mono-synaptic pathway to develop separable invertible pathways where a given EC input pattern can be encoded over a sparser representation in the corresponding CA1 slot, and this CA1 representation can in turn recover the full original EC slot pattern.

The learning dynamics within this model, as governed by the LEABRA algorithm, are dictated by the phases of a presumed theta oscillation, thus its shot-hand name of 'Theta-Phase' used here. In short, during learning two error signals are derived within every period of a give theta oscillation.  The first, during the trough of the theta oscillation, provides an error driven learning signal to the mono-synaptic pathway, learning the invertible mapping between EC and CA1 can be thought of as an encoding.  The second, during the peak of the theta oscillation, provides a error driven learning signal on the tri-synaptic pathway, learning to recollect missing or incomplete information can be thought of as retrieval.

Building on this model that inherently captures the hippocampal theta dynamics, we seek to add a modulatory attention signal that can capture the appropriate alpha dynamics derived from the thalamo-cortical principles described above.  This following section reviews the specifics of that implementation, and the testing paradigm used to validate its performance.

\subsection{Thalamic Modulation as Alpha}
\label{sec:attmod}
The effective implementation of the selective attention mechanism into the Theta-Phase model is done through a separate layer that reflects the thalamic relay cell (TRC) activity, as well as an attentional modulation value calculated from this TRC activity that scales the net input of downstream units.  This can be seen implemented in the Theta-Phase model in Figure \ref{fig:depNet}.  Here connectivity between the posterior cortical inputs and Entorhinal Cortex (EC) also goes through a `TRC' layer corresponding to the thalamic relay cells that then supply the attentional modulation signal.  Each of the posterior cortical regions maintains topographic input into EC and the corresponding TRC, which is manifest in the model as separate unit groups as shown in Figure \ref{fig:depNet}B.  Each of these unit groups has a single TRC unit that calculates its modulatory signal as determined by the activation in the corresponding posterior cortical region and EC unit group.  Finally, a single modulatory scalar value is calculated for each of the TRCs and sent to its corresponding EC unit group either inhibiting activation in that unit group or allowing normal processing to occur.

The detailed description of this modulator mechanism's implementation is described in supplementary info section \ref{appDeepLeabra}, however a coarse description and its functional implications are described here.  The attentional modulation is effectively a normalized scaling factor based on the net input to a given TRC unit.  A given TRC unit receives inputs with fixed weighted connections (i.e. no learning signal adjusts these weights), and its activation is used to calculate a single value that multiplicitively scales the net input of each of the units it is connected to.  This attentional scaling value is simply the TRC activation normalized to range from some pre-determined minimum to a maximum of 1.  This pre-determined minimum scaling factor can be thought of as the most inhibitory influence possible, and was loosely fit for these simulations to a value of 0.25 to match the anticipated accuracies across conditions.  It should be noted this minimum scaling value did not vary across trials or conditions and therefore is unable to bias reactivation.

\subsection{Model Validation}
\label{sec:ModelTest}

The Dependent Events paradigm (see Figure \ref{fig:deev}) was chosen to test the performance of the combined Theta-Phase model with attentional gating.  This paradigm ideally captures behavioral components related to Complimentary Learning Systems ideas of pattern completion and separation, as well as components relating to selective attention over multi-element memory traces.  These two aspects allow for the testing of the computational model to match known behavioral performance and relate the findings to predictions in the EEG time frequency domain.

Basing the testing procedure off \textcite{HornerBisbyBushEtAl15}, the paradigm was implemented as close to the behavioral version as possible.  The model learned a series of multi-element `events', see Figure \ref{fig:deev}. Each event consisted of three or four elements (locations, people, objects and animals), plus 4 contextual elements that were consistent across presentations. Events were built up over three separate encoding trials (interleaved with encoding trials of other events). Each trial consisted of the presentation of one of the three possible pairwise associations from a given event. This paradigm allows for the building of `events' with different associative structures of overlapping pairs: `closed-loop', in which all event elements were presented paired with all other elements of the event; or `open-loop', in which elements of an event were presented as a chain of overlapping pairs. Here, the term `event' groups  the set of overlapping associations encoded across separate paired associate trials.

The model's input patterns, described in more detail Section \ref{sec:HippModelValidation}, were created with a base vocabulary of 100 7x7 unit group patterns with 7 units active and a minimum Hamming distance of 10 between each pattern.  The composite input patterns used to train the model were generated from this set of vocabulary patterns, with 8 unit groups in each input pattern.  These input patterns were conceptualized to have 4 unit groups that each corresponded to one of the various element types of the paired associates used in the Dependent Events paradigm.  For each event, one unit group had a unique pattern of activity that corresponded to `location' representations, another for `people', another for `objects', and another for `animals'.  These patterns were consistent across presentations of a given event, i.e. the same unit group patterns were used across the 3 encoding trials, and 6 retrieval trials for each event. In this way selective attention during encoding and retrieval can be  guided to these specific stimulus dimensions of the input pattern, as shown in Figure \ref{fig:depNet}C where two element types are being presented and the corresponding attentional modulation signals are being shown inhibiting the non-attended dimensions.  In addition to these 4 unit groups corresponding to the different event element types, 4 incidental unit groups also maintained the same representation across event presentations.  This was conceptualized as other contextual elements of the event participants imagined, and likely aided in the model's ability to recall the associations (although simulations without these contextual elements were not explored).

\subsection{Testing Procedure}
\label{sec:TestProc}
The encoding phase was split into three mini-blocks. One pairwise association for each of the 36 events was presented during each mini-block. Presentation order within each mini-block was randomized. For the closed-loop events, the order of presentation across the mini-blocks for a specific event was either: (1) location-object/animal, location-person and person-object/animal, (2) person-object/animal, location-person and location-object/animal, (3) location-person, location-object/ animal, person-object/animal or (4) location-person, person-object/animal and location-object/animal. For open-loop events, it was either: (1) location-object, location-person, person-animal, (2) person-animal, location-person, location-object, (3) location-person, location-object, person-animal or (4) location-person, person-animal, location-object. Each of the four orderings was run on two random initialization of weights, to make up the pool of results. The model was trained on 2 epochs (random permutations of all the trials within a mini-block) for each mini-block before starting training on the next mini-block.  Thus, for the first and second mini-blocks, the closed-loop and open-loop conditions were identical, presenting a single pairwise association and then a second overlapping pairwise association. %The third encoding trial either formed an all-to-all associative structure in the closed-loop condition or an associative chain in the open-loop condition (Fig. 1).

At retrieval, the model was tested on each encoded pairwise association for each event in both directions.  In this way, each event was tested across six separate retrieval trials.  In the behavioral paradigm each retrieval trial was a six-forced choice recognition test with a particular cue element type and potential 6 targets all from the same element type (for example, cue location, retrieve the associated person among five people from other events).  However, in the model, this was captured by providing an activation of 0.1429 to all units in the target element dimension, providing a similar net activation compared to the complete patterns (i.e. $49 * 0.1429=7$), and thus opening the thalamic gate to allow processing in the target dimension. Note that a single cue pattern, and non-informative target dimension activation were presented on all retrieval trials with the task being to reactivate the original paired associate.  This reactivation success was measured by the binary Name Error metric described in Section \ref{sec:HippModelValidation}, i.e. either correct reactivation or not for a given cue-target pair. Thus, as with encoding, the closed-loop and open-loop conditions were exactly matched in terms of stimuli and task demands, differing only in the potential occurrence of incidental reactivation (that is, different levels of pattern completion).  An example of what this looks like in the network is shown in Figure \ref{fig:netNonTarg}B, with more pattern completion occurring in the location stimulus dimension (both in ECout and the TRC layer) for the closed-loop condition compared to open-loop.

Two parameters were loosely fit to match retrieval performance from previous behavioral results of the Dependent Events paradigm \cite{HornerBisbyBushEtAl15,HornerBurgess14}.  The first was the number of training repetitions (referred to as `epochs') for each mini-block of encoding.  The second was the minimum amount of attentional modulation, i.e. when the thalamic gate is fully closed how much activation is allow to get through, see Section \ref{sec:attmod} and Appendix \ref{appDeepLeabra} for more details.  By coarsely comparing results from Name Error determined retrieval accuracy in the network, shown in Figure \ref{fig:netDep}, to the known behavioral results from \textcite{HornerBisbyBushEtAl15,HornerBurgess14}, the two parameters were fit (2 epochs of training per mini-block, and a minimum activation of 0.25) to loosely match network performance with human behavior.

\subsection{Measure of Event Dependency} 
\label{sec:CalcDep}
Based on \textcite{HornerBisbyBushEtAl15}, dependency between event elements was assessed by constructing contingency tables for successfully retrieving, on separate trials, the two associated elements (for example, person and object) when cued by the remaining element (for example, location), as well as for retrieving a particular target element (for example, location) when cued by the other two elements (for example, person and object), see Table \ref{tab:depEvt} for example contingency tables. Averaging across events within conditions, this resulted in six 2x2 tables per model run per condition with the on-diagonal elements reflecting the Dependency for a given cue or target element.  These were then averaged across the 4 contingency tables with shared associations across both open and closed loop conditions (i.e. location and object cue and target tables) to get a single measure of dependency per condition.

To accurately compare condition differences in Dependency between open and closed-loop events its necessary to control for accuracy differences.  For example, if all paired associates are correctly retrieved it's unclear if there is inherent Dependency between event elements or if each association was memorized independently.   To address this, an `Independent' model that reflects the expected Dependency irrespective of the underlying event structure can be calculated to compare with the Dependency inherent in the event structures. For a given model random weight initialization, the mean proportion of correct retrievals for a given element type, say $B$, (over all events) when cued by a given element type, $A$ , can be denoted as $P_{AB}$. This accuracy can be used to create contingency tables for an Independent model where the probability of correctly or incorrectly retrieving two associations is simply the product of the mean probability for each association (i.e., the proportion of events for which both $B$ and $C$ were correctly retrieved when cued by $A$ would be $P_{AB}  P_{AC}$, while the proportion where $B$ was correctly retrieved but $C$ was not would be $P_{AB}(1-P_{AC})$).  A example contingency tables for the Independent model are show in Table \ref{tab:depInd}.

\begin{table}
\parbox{.45\linewidth}{
\centering
\begin{tabular}{l | c c}
	Cue$_{A_i}$ & Hit$_{C_i}$ & Miss$_{C_i}$ \\
	\hline
	Hit$_{B_i}$ & $H_{A_i B_i} \land H_{A_i C_i}$ & $H_{A_i B_i} \land M_{A_i C_i}$\\
	Miss$_{B_i}$ & $M_{A_i B_i} \land H_{A_i C_i}$ & $M_{A_i B_i} \land M_{A_i C_i}$\\	
	\hline
\end{tabular}
\newline
\vspace{5mm}
\newline
\begin{tabular}{l | c c}
	Targ$_{A_i}$ & Hit$_{C_i}$ & Miss$_{C_i}$ \\
	\hline
	Hit$_{B_i}$ & $H_{B_i A_i} \land H_{C_i A_i}$ & $H_{B_i A_i} \land M_{C_i A_i}$\\
	Miss$_{B_i}$ & $M_{B_i A_i} \land H_{C_i A_i}$ & $M_{B_i A_i} \land M_{C_i A_i}$\\	
	\hline
\end{tabular}
%\vspace{5mm}
\caption[Contingency Tables for Data driven Dependency]{Contingency tables for a given event $i$ focused on element $A$.  $H_{A_i B_i }$ implies a correct retrieval with cue element $A$ and target element $B$ from event $i$, while $M_{A_i B_i}$ implies an incorrect retrieval with the same cue and target, $\land$ implies a logical and. \label{tab:depEvt}
}
}
\hfill
\parbox{.45\linewidth}{
\centering
\begin{tabular}{l|cc}
	Cue$_{A}$ & Hit$_{C}$ & Miss$_{C}$ \\
	\hline
	Hit$_{B}$ & $P_{AB} P_{AC}$ & $P_{AB} (1-P_{AC})$ \\
	Miss$_{B}$ & $(1-P_{AB}) P_{AC}$ & $(1-P_{AB}) (1-P_{AC})$ \\	
	\hline
\end{tabular}
\newline \vspace{5mm} \newline
\begin{tabular}{l|cc}
	Targ$_{A}$ & Hit$_{C}$ & Miss$_{C}$ \\
	\hline
	Hit$_{B}$ & $P_{BA} P_{CA}$ & $P_{BA} (1-P_{CA})$ \\
	Miss$_{B}$ & $(1-P_{BA}) P_{CA}$ & $(1-P_{BA}) (1-P_{CA})$ \\	
	\hline
\end{tabular}
\caption[Contingency Tables for Independent Dependency]{Independent model contingency table focused on element $A$ across all events.  $P_{AB}$ implies percent correct across all retrieval trials where element category $A$ is the cue and element category $B$ is the target. \label{tab:depInd} 
}
}
\end{table}

\section{Behavioral and EEG Methods}\label{sec:behmethods}

\subsection{Participants}
49 University of Colorado undergraduates participated in the experiment and received payment of \$15 per hour (ages 18-29, mean$=$19.76; 30 male, 19 female). All participants were right handed, had normal or corrected-to-normal vision, and all but 5 were native english speakers.  Informed consent was obtained from each participant, and the study conformed to the Human Research Committee guidelines.  Two participants were removed due to equipment malfunction, nineteen were removed due excessively noisy EEG data such that they had less then 20 trials in any experimental condition.  This excessive number of participants removed based on low trial counts is likely due to two factors: first is the paradigm requires participants to move their eyes, causing more EEG artifacts then would be expected otherwise, and second is the sub-selection of trials by correctness and confidence as discussed in Section \ref{sec:deevEEGConf}.

\subsection{Materials}
Experimental stimuli, originally adapted from \textcite{HornerBisbyBushEtAl15}, were 252 individual words split into 4 categories with 72 locations, 72 people, 54 objects, and 54 animals.  72 individual `events' were generated for each participant, half of which were closed-loop events consisting of three elements (e.g. location-person-object), and half were open-loop events consisting of 4 elements.  Half of the closed-loop events consisted of location-person-object triplets and half consisted of location-person-animal triplets.  These 72 events were equally split into two blocks balancing across all event types, and the subset of words used to generate events in block 1 was the same across participants.  An example set of events for block 1 is shown in Appendix \ref{appStimuli}.

>>>>

@@

%add 'S' prefix to figure and section 
\renewcommand{\thefigure}{S\arabic{figure}}
\renewcommand{\thesection}{S\arabic{section}}

\begin{document}
